### 2018-02-07 Maths Notes/Statistics/S3/Estimation/Estimation.tex
Notes Update

\begin{document}
\begin{center}
\underline{\huge Estimation, confidence intervals and tests}
\end{center}
\section{Estimators}
A statistic that is used to estimate a population parameter is an \textbf{estimator}.\\
A particular individual value is an \textbf{estimate}.\\
\textbf{Bias} is how far the estimator is from the true value\\
\\
If a statistic T is used as an estimator for a population parameter $\theta$ then the bias is:
$$E(T)-\theta$$
If $E(T)=\theta$ then the statistic is unbiased
\subsection{Proving $\mathbf{E(\overline{X})=\mu}$}
\textit{Prove that $\overline{X}$ is an unbiased estimator for $\mu$ when the population is normally distributed}\\
\\
A random sample $X_1,X_2,X_3...X_N$ is taken for a population with $X\sim N(\mu,\sigma^2)$
$$\overline{X}=\frac{1}{n}\sum X$$
$$E(\overline{X})=E\Big(\frac{1}{n}\sum X\Big)$$
$$E(\overline{X})=\frac{1}{n}\big(E(X_1)+E(X_2)+E(X_3)...+E(X_N)\big)$$
$$E(\overline{X})=\frac{1}{n}(\mu+\mu+\mu...+\mu)$$
$$E(\overline{X})=\frac{1}{n}(n\mu)$$
$$E(\overline{X})=\mu$$
\subsection{Variance}
$$Var(\overline{X})=\frac{\sigma^2}{\textrm{Sample size}}$$
\\
We use $S^2$ as an estimator for $\sigma^2$
$$S^2=\frac{1}{n-1}\Big(\Sigma X^2-n\overline{X}^2\Big)$$
\subsection{Uniform Distributions}
\textit{Random variable X is continuously uniform $[0,\alpha]$. A sample $X_1,X_2..X_N$ is taken}\\
\textit{Show that $\overline{X}$ is a biased estimate and state the bias}
$$\overline{X}=\frac{0+\alpha}{2}=\frac{\alpha}{2}$$
$$E(\overline{X})=E(\frac{\alpha}{2})=\frac{1}{2}E(\alpha)=\frac{\alpha}{2}$$
$$\textrm{Bias}=\frac{\alpha}{2}-\alpha=-\frac{\alpha}{2}$$
\newpage
\section{Standard error}
$$\textrm{Standard error}=\frac{\sigma}{\sqrt{n}} \ \textrm{or} \ \frac{s}{\sqrt{n}}$$
\section{Central limit theorem}
The central limit theorem (C.L.T.) states that if $X_1,X_2,X_3...$ is a random sample, from any distribution with mean $\mu$ and variance $\sigma^2$. Then the sample means, $\overline{X}$ are distributed with a normal distribution.
$$\overline{X}\sim N(\mu,\frac{\sigma^2}{n})$$ 
For this, the sample size must be greater than 50
\section{Confidence intervals}
A 95\% confidence interval is two values in which there is a probability of 0.95 that it will contain the population mean.\\
\begin{tikzpicture}
\begin{axis}[
  no markers, domain=0:8, samples=100,
  every axis y label/.style={at=(current axis.above origin),anchor=south},
  every axis x label/.style={at=(current axis.right of origin),anchor=west},
  height=5cm, width=12cm,
  xtick={2,6},xticklabels={0.025,0.025}, ytick=\empty,
  enlargelimits=false, clip=false,   ]
  \addplot [fill=cyan!20, draw=none, domain=0:2] {gauss(4,1)} \closedcycle;
  \addplot [fill=cyan!20, draw=none, domain=6:8] {gauss(4,1)} \closedcycle;
  \addplot [very thick,cyan!50!black] {gauss(4,1)};
  \node[] at (axis cs: 4,0.1) {95\%};
\end{axis}
\end{tikzpicture}
\\
Look up the value of one of the tails on the percentage points table and substitute into the formula.\\
For example:\\
At a 95\% confidence interval each of the tails is 0.025 and the corresponding z value is 1.96.\\
$$\overline{x}\pm1.96\times\frac{\sigma}{\sqrt{n}}$$

>>>>

\begin{document}
\begin{center}
\underline{\huge Estimation, confidence intervals and tests}
\end{center}
\section{Estimators}
A statistic that is used to estimate a population parameter is an \textbf{estimator}.\\
A particular individual value is an \textbf{estimate}.\\
\textbf{Bias} is how far the estimator is from the true value\\
\\
If a statistic T is used as an estimator for a population parameter $\theta$ then the bias is:
$$E(T)-\theta$$
If $E(T)=\theta$ then the statistic is unbiased
\subsection{Proving $\mathbf{E(\overline{X})=\mu}$}
\textit{Prove that $\overline{X}$ is an unbiased estimator for $\mu$ when the population is normally distributed}\\
\\
A random sample $X_1,X_2,X_3...X_N$ is taken for a population with $X\sim N(\mu,\sigma^2)$
$$\overline{X}=\frac{1}{n}\sum X$$
$$E(\overline{X})=E\Big(\frac{1}{n}\sum X\Big)$$
$$E(\overline{X})=\frac{1}{n}\big(E(X_1)+E(X_2)+E(X_3)...+E(X_N)\big)$$
$$E(\overline{X})=\frac{1}{n}(\mu+\mu+\mu...+\mu)$$
$$E(\overline{X})=\frac{1}{n}(n\mu)$$
$$E(\overline{X})=\mu$$
\subsection{Variance}
$$Var(\overline{X})=\frac{\sigma^2}{\textrm{Sample size}}$$
\\
We use $S^2$ as an estimator for $\sigma^2$
$$S^2=\frac{1}{n-1}\Big(\Sigma X^2-n\overline{X}^2\Big)$$
\subsection{Uniform Distributions}
\textit{Random variable X is continuously uniform $[0,\alpha]$. A sample $X_1,X_2..X_N$ is taken}\\
\textit{Show that $\overline{X}$ is a biased estimate and state the bias}
$$\overline{X}=\frac{0+\alpha}{2}=\frac{\alpha}{2}$$
$$E(\overline{X})=E(\frac{\alpha}{2})=\frac{1}{2}E(\alpha)=\frac{\alpha}{2}$$
$$\textrm{Bias}=\frac{\alpha}{2}-\alpha=-\frac{\alpha}{2}$$
\newpage
\section{Standard error}
$$\textrm{Standard error}=\frac{\sigma}{\sqrt{n}} \ \textrm{or} \ \frac{s}{\sqrt{n}}$$
\section{Central limit theorem}
The central limit theorem (C.L.T.) states that if $X_1,X_2,X_3...$ is a random sample, from any distribution with mean $\mu$ and variance $\sigma^2$. Then the sample means, $\overline{X}$ are distributed with a normal distribution.
$$\overline{X}\sim N(\mu,\frac{\sigma^2}{n})$$ 
For this, the sample size must be greater than 50
\section{Confidence intervals}
A 95\% confidence interval is two values in which there is a probability of 0.95 that it will contain the population mean.\\
\begin{tikzpicture}
\begin{axis}[
  no markers, domain=0:8, samples=100,
  every axis y label/.style={at=(current axis.above origin),anchor=south},
  every axis x label/.style={at=(current axis.right of origin),anchor=west},
  height=5cm, width=12cm,
  xtick={2,6},xticklabels={0.025,0.025}, ytick=\empty,
  enlargelimits=false, clip=false,   ]
  \addplot [fill=cyan!20, draw=none, domain=0:2] {gauss(4,1)} \closedcycle;
  \addplot [fill=cyan!20, draw=none, domain=6:8] {gauss(4,1)} \closedcycle;
  \addplot [very thick,cyan!50!black] {gauss(4,1)};
  \node[] at (axis cs: 4,0.1) {95\%};
\end{axis}
\end{tikzpicture}
\\
Look up the value of one of the tails on the percentage points table and substitute into the formula.\\
For example:\\
At a 95\% confidence interval each of the tails is 0.025 and the corresponding z value is 1.96.\\
$$\overline{x}\pm1.96\times\frac{\sigma}{\sqrt{n}}$$
\section{Hypothesis testing}
The test statistic for the population mean $\mu$ is $Z=\dfrac{X-\mu}{\frac{\sigma}{\sqrt{n}}}$\\
\\
A hypothesis test can be done quicker by finding the critical values from the percentage points table.
\subsection{Example}
$\mu=0.58 \ \sigma=0.015$\\
Measured\\
$n=50 \ \mu=0.577 \ \sigma=0.015$\\
\\
$H_0: \mu=0.50$\\
$H_1: \mu\neq0.58$\\
$$\overline{D}\sim N(0.58,\frac{0.015^2}{50})$$
$$Z=\frac{0.577-0.58}{\sqrt{\frac{0.015^2}{50}}}=-1.41$$
At 0.5\%  per level Z=2.5758
$$-1.41>-2.5758$$
Not in critical region so not significant, no evidence to reject $H_0$, the diameter of the bolts is not changed.

### 2018-02-07 Physics Notes/Nuclear Physics/Radioactivity/Radioactivity.tex
Notes Update

\includegraphics[width=6cm]{foil.jpg}  
    \end{minipage}\hfill
    \begin{minipage}{0.45\textwidth}
        \centering
\begin{tabularx}{\textwidth}{|X|X|}
\hline
Observation&Explanation\\
\hline
Most electrons pass all the way through&Atoms are mostly empty space\\
\hline
Some are deflected&The atom has a positive centre\\
\hline
Some are deflected by significant angles&The positive charge is condensed in a small area\\
\hline
\end{tabularx}
    \end{minipage}
\end{figure}
\subsection{Estimating the size of the nucleus}  
\subsubsection{Closest approach method}
\begin{tikzpicture}
\filldraw[color=red, fill=red, very thick](0,0) circle (0.2);
\draw[thick, ->,black] (0.4,0) -- (4,0) node[above,yshift=0.3cm,xshift=-2cm] {\footnotesize{Velocity reduces on approach}};
\draw[color=red, very thick](6,0) circle (1) node[] {Au};
\draw[color=black, very thick,dashed](6,0) circle (2) node[above,yshift=2cm] {EPE};
\draw[thick, ->,black] (4,0) -- (1,-1);
\end{tikzpicture}\\
\\
KE=EPE
$$8.0\times10^{-13}=\frac{1}{4\pi\epsilon_0}\times\frac{Q_{Au}}{r}\times Q_{\alpha}$$
$$r=4.55\times10^{-14}$$
\subsubsection{Estimate from scattering data}
\begin{itemize}
\item About $\frac{1}{10,000}$ deflected through more than 90\degree
\item Foil had n layers of atoms
\end{itemize}
\begin{tikzpicture}
\draw[thick, stealth-stealth,black] (6.8,-0.4) -- (6.8,0.4) node[right,midway] {d} ;
\draw[color=red, very thick](6,0) circle (0.5) node[] {\tiny{Nucleus}};
\draw[color=black, very thick](6,0) circle (2) node[above,yshift=2cm] {Atom};
\draw[thick, stealth-stealth,black] (8.5,-2) -- (8.5,2) node[right,midway] {D};
\end{tikzpicture}\\
$$\frac{\frac{1}{4}\pi d^2}{\frac{1}{4}\pi D^2}=\frac{d^2}{D^2}=\frac{1}{10,000n}$$
$n=10^4$ layers
$$\frac{d^2}{D^2}=\frac{1}{10,000\times1\times10^4}$$
$$d=\frac{D}{10,000}$$
\newpage
\section{Radioactive materials}
\subsection{Sources of background radiation by most common}
\begin{enumerate}
\item Air (e.g. radon gas)
\item Medical
\item Ground and buildings
\item Food and drink
\item Cosmic rays
\item Nuclear weapons
\item Air travel
\item Nuclear power
\end{enumerate}
\subsection{Geiger M\"{u}ller tube}
\includegraphics[width=10cm]{geiger.png}\\
When a particle of ionising radiation enters the tube, the particle ionises the gas atoms along its track. The negative ions are attracted to the rod and the positive ions to the wall. These ions cause further ionisation, creating enough ions for a current to flow. A pulse of charge passes round the circuit through resistor R, causing the voltage pulse across R which is recorded as a single count by the pulse counter\\
\\
The dead time of the tube, the time taken to regain its non conducting state after an ionising particle enters it, is typically of the order of 0.2ms.
\newpage
\begin{tabularx}{\textwidth}{|c|X|X|X|}
\hline
&Alpha&Beta&Gamma\\
\hline
Nature&2 Protons+2 Neutrons&High speed electron or positron&High energy photon\\
\hline
Range&Up to 10cm&Up to 1m&Infinite\\
\hline 
Deflection in a magnetic field&Deflected&Opposite direction to $\alpha$ particles and more easily deflected&Not deflected\\
\hline
Absorption&Paper&Aluminium&Lead \\
\hline
Ionisation&$10^4$ ions per mm&100 ions per mm&Very weak ionising effect\\
\hline
Energy of each particle&Constant for a given source&Varies up to a maximum for a given source&Constant for a given source\\
\hline
\end{tabularx}

>>>>

\includegraphics[width=6cm]{foil.jpg}  
    \end{minipage}\hfill
    \begin{minipage}{0.45\textwidth}
        \centering
\begin{tabularx}{\textwidth}{|X|X|}
\hline
Observation&Explanation\\
\hline
Most electrons pass all the way through&Atoms are mostly empty space\\
\hline
Some are deflected&The atom has a positive centre\\
\hline
Some are deflected by significant angles&The positive charge is condensed in a small area\\
\hline
\end{tabularx}
    \end{minipage}
\end{figure}
\subsection{Estimating the size of the nucleus}  
\subsubsection{Closest approach method}
\begin{tikzpicture}
\filldraw[color=red, fill=red, very thick](0,0) circle (0.2);
\draw[thick, ->,black] (0.4,0) -- (4,0) node[above,yshift=0.3cm,xshift=-2cm] {\footnotesize{Velocity reduces on approach}};
\draw[color=red, very thick](6,0) circle (1) node[] {Au};
\draw[color=black, very thick,dashed](6,0) circle (2) node[above,yshift=2cm] {EPE};
\draw[thick, ->,black] (4,0) -- (1,-1);
\end{tikzpicture}\\
\\
KE=EPE
$$8.0\times10^{-13}=\frac{1}{4\pi\epsilon_0}\times\frac{Q_{Au}}{r}\times Q_{\alpha}$$
$$r=4.55\times10^{-14}$$
\subsubsection{Estimate from scattering data}
\begin{itemize}
\item About $\frac{1}{10,000}$ deflected through more than 90\degree
\item Foil had n layers of atoms
\end{itemize}
\begin{tikzpicture}
\draw[thick, stealth-stealth,black] (6.8,-0.4) -- (6.8,0.4) node[right,midway] {d} ;
\draw[color=red, very thick](6,0) circle (0.5) node[] {\tiny{Nucleus}};
\draw[color=black, very thick](6,0) circle (2) node[above,yshift=2cm] {Atom};
\draw[thick, stealth-stealth,black] (8.5,-2) -- (8.5,2) node[right,midway] {D};
\end{tikzpicture}\\
$$\frac{\frac{1}{4}\pi d^2}{\frac{1}{4}\pi D^2}=\frac{d^2}{D^2}=\frac{1}{10,000n}$$
$n=10^4$ layers
$$\frac{d^2}{D^2}=\frac{1}{10,000\times1\times10^4}$$
$$d=\frac{D}{10,000}$$
\newpage
\section{Radioactive materials}
\subsection{Sources of background radiation by most common}
\begin{enumerate}
\item Air (e.g. radon gas)
\item Medical
\item Ground and buildings
\item Food and drink
\item Cosmic rays
\item Nuclear weapons
\item Air travel
\item Nuclear power
\end{enumerate}
\subsection{Geiger M\"{u}ller tube}
\includegraphics[width=10cm]{geiger.png}\\
When a particle of ionising radiation enters the tube, the particle ionises the gas atoms along its track. The negative ions are attracted to the rod and the positive ions to the wall. These ions cause further ionisation, creating enough ions for a current to flow. A pulse of charge passes round the circuit through resistor R, causing the voltage pulse across R which is recorded as a single count by the pulse counter\\
\\
The dead time of the tube, the time taken to regain its non conducting state after an ionising particle enters it, is typically of the order of 0.2ms.\\
\section{Radioactive decay}
\begin{tabularx}{\textwidth}{|c|X|X|X|}
\hline
&Alpha&Beta&Gamma\\
\hline
Nature&2 Protons+2 Neutrons&High speed electron or positron&High energy photon\\
\hline
Range&Up to 10cm&Up to 1m&Infinite\\
\hline 
Deflection in a magnetic field&Deflected&Opposite direction to $\alpha$ particles and more easily deflected&Not deflected\\
\hline
Absorption&Paper&Aluminium&Lead \\
\hline
Ionisation&$10^4$ ions per mm&100 ions per mm&Very weak ionising effect\\
\hline
Energy of each particle&Constant for a given source&Varies up to a maximum for a given source&Constant for a given source\\
\hline
\end{tabularx}
\newpage
\subsection{$\alpha$ decay}
